---
title: "Fixed versus Random Effects Meta-analysis"
author: Daniel Noble
date: '`r Sys.Date()`'
bibliography: ./bib/refs.bib
csl: ./bib/the-journal-of-experimental-biology.csl
output:
  bookdown::html_document2:
    css: style.css
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, tidy = TRUE)
options(digits=2)
```

```{r klippy, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
#install.packages("devtools")
remotes::install_github("rlesur/klippy")
klippy::klippy(tooltip_message = 'Click to Copy Code', tooltip_success = 'Done', position = 'right', color = "red")

# Load packages
pacman::p_load(metafor, flextable, tidyverse, orchaRd, pander, mathjaxr, equatags, vembedr)

# To use mathjaxr 
```

## What's the difference between fixed and random effects meta-analysis?

To demonstrate, we can simulate some data. This is also quite useful for getting a handle on what the difference between fixed and random effects meta-analytic models looks like. Either way, we'll explore the data in depth as we move through the tutorial. 

```{r, simdat, eval=TRUE, echo=TRUE}
# Generate some simulated effect size data with known sampling variance
# assumed to come from a common underlying distribution
set.seed(86) # Set see so that we all get the same simulated results
 # We will have 5 studies
    stdy  <- 1:5                                      
# We know the variance for each effect
    Ves     <- c(0.05, 0.10, 0.02, 0.10, 0.09)    
# We'll need this later but these are weights
    W     <- 1 / Ves                                          
# We assume they are sampled from a normal distribution  with a mean effect size of 2
    es     <- rnorm(length(Ves), 2, sqrt(Ves))          
# Data for our fixed effect meta-analysis
    dataFE <- data.frame(stdy = stdy, es, Ves)

# Generate a second set of effect sizes, but now assume that each study effect does not come from the same distribution, but from a population of effect sizes. 

# Here adding 0.8 says we want to add 0.8 as the between study variability. In other words, each effect size is sampled from a larger distribution of effect sizes that itself comes from a distribution with a variance of 0.8. 
    esRE        <- rnorm(length(Ves), 2, sqrt(Ves + 0.8)) 
# Data for our random effect meta-analysis 
    dataRE <-  data.frame(stdy = stdy, esRE, Ves)
```

We can get a look at what these two datasets we simulated above look like (Figure \@ref(fig:fvsr)). The red circles are effect sizes and their standard errors (square root of the sampling error variance) from our fixed effect meta-analysis (FE) data set. In contrast, the black circles are our effect size and standard errors from the data generated in the random effect meta-analysis (RE) dataset. The black line is the average, true effect size (which we have set to the value 2).

```{r, fvsr, eval=TRUE, echo=FALSE, fig.cap="Mean (arrows are sampling standard deviation) effect size for each study. Data simulated under a fixed effect model in black and data simulated under a random effect model in red."}
    
  plot(stdy - 0.25 ~ es, col = "black", pch = 16, ylim = c(0, 6), xlim = c(1, 3.2), 
       data = dataFE, ylab = "Study", xlab = "Effect size")
  abline(v = 2, lty =3)
  arrows(x0 = dataFE$es, y0 = dataFE$stdy- 0.25, y1 = dataFE$stdy- 0.25, x1 = dataFE$es + sqrt(dataFE$Ves), 
         angle = 90, length = 0.05, col = "black")
  arrows(x0 = dataFE$es, y0 = dataFE$stdy- 0.25, y1 = dataFE$stdy- 0.25, x1 = dataFE$es - sqrt(dataFE$Ves), 
         angle = 90, length = 0.05, col = "black")
  
  points(stdy + 0.25 ~ esRE, col = "red", pch = 16, data = dataRE)
  arrows(x0 = dataRE$es, y0 = dataRE$stdy+ 0.25, y1 = dataRE$stdy+ 0.25, x1 = dataRE$es + sqrt(dataRE$Ves), 
         angle = 90, length = 0.05, col = "red")
  arrows(x0 = dataRE$es, y0 = dataRE$stdy+ 0.25, y1 = dataRE$stdy+ 0.25, x1 = dataRE$es - sqrt(dataRE$Ves), 
         angle = 90, length = 0.05, col = "red")
  
```

We notice a few important differences here. In the RE dataset the variance across the studies is a lot larger compared to the FE dataset. This is what we would expect because we have added in a between study variance. Now, let’s use a common meta-analysis package, *metafor*, to analyse these datasets.

```{r, fixedma,  eval=TRUE, echo=TRUE}
# Run a fixed effect meta-analysis using the FE dataset. 
    metafor::rma(yi = es, vi = Ves, method = "FE", data = dataFE)
```

So this suggests that the average estimate is 2.07 and it has a standard error of 0.1. It also provides us with a Q statistic, which measures the amount of heterogeneity among effect sizes. If Q is large and the p-value is small then it suggests substantial heterogeneity among effect sizes beyond what we would expect if these effects were generated from a common underlying distribution. This is a major assumption of fixed effect meta-analyses and in this case is supported. This is a good thing because we specifically generated these data under the assumption that they are from the same distribution. We can see this if we look at how we generated the data: `rnorm(length(Ves), 2, sqrt(Ves))`. This draws random effect sizes from a normal distribution with a variance (`sqrt(Ves)`) for each effect size that is only defined by its sampling variability.

What is this model doing though and what is the logic behind the calculations? The best way to understand this is to hand calculate these values. Basically the effect sizes are being weighted by their sampling error variance when deriving the pooled estimate and it’s variance. Let’s calculate this by hand and see what’s happening:

```{r, fixedmahand,  eval=TRUE, echo=TRUE}
# Calculate pooled effect size
       EsP.FE <- sum(W*dataFE$es) / sum(W)
       EsP.FE
# Calculate the pooled variance around estimate
    VarEsP.FE <- 1 / sum(W)
    VarEsP.FE
# Calculate the standard error around estimate
    SE.EsP.FE <- sqrt(VarEsP.FE)
    SE.EsP.FE
```

Wow! This is so cool. We just did a fixed effect meta-analysis by hand…and, look, it matches the model output perfectly. The math is not so scary after all! But what about this extra statistic, Q? How do we derive this?

```{r, q_fe,  eval=TRUE, echo=TRUE}
#Now lets calculate our Q. Q is the total amount of heterogeneity in our data.
    Q.fe <- sum(W*(dataFE$es^2) ) - (sum(W*dataFE$es)^2 / sum(W))
    Q.fe
```

Cool. So this value also matches up nicely. Now lets move on to a more complicated situation, a random effect meta-analysis using the RE dataset. Remember, we know from this data set that each effect size comes from a different overall distribution. How might Q change? (Hint: We expect it to get larger!)
