---
title: "Fixed versus Random Effects Meta-analysis"
author: Daniel Noble
date: '`r Sys.Date()`'
bibliography: ./bib/refs.bib
csl: ./bib/the-journal-of-experimental-biology.csl
output:
  bookdown::html_document2:
    css: style.css
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---



## What's the difference between fixed and random effects meta-analysis?

To demonstrate, we can simulate some data. This is also quite useful for getting a handle on what the difference between fixed and random effects meta-analytic models looks like. Either way, we'll explore the data in depth as we move through the tutorial. 

```{r, simdat, eval=TRUE, echo=TRUE}
# Generate some simulated effect size data with known sampling variance
# assumed to come from a common underlying distribution
set.seed(86) # Set see so that we all get the same simulated results
 # We will have 5 studies
    stdy  <- 1:5                                      
# We know the variance for each effect
    Ves     <- c(0.05, 0.10, 0.02, 0.10, 0.09)    
# We'll need this later but these are weights
    W     <- 1 / Ves                                          
# We assume they are sampled from a normal distribution  with a mean effect size of 2
    es     <- rnorm(length(Ves), 2, sqrt(Ves))          
# Data for our fixed effect meta-analysis
    dataFE <- data.frame(stdy = stdy, es, Ves)

# Generate a second set of effect sizes, but now assume that each study effect does not come from the same distribution, but from a population of effect sizes. 

# Here adding 0.8 says we want to add 0.8 as the between study variability. In other words, each effect size is sampled from a larger distribution of effect sizes that itself comes from a distribution with a variance of 0.8. 
    esRE        <- rnorm(length(Ves), 2, sqrt(Ves + 0.8)) 
# Data for our random effect meta-analysis 
    dataRE <-  data.frame(stdy = stdy, esRE, Ves)
```

We can get a look at what these two datasets we simulated above look like (Figure \@ref(fig:fvsr)). The red circles are effect sizes and their standard errors (square root of the sampling error variance) from our fixed effect meta-analysis (FE) data set. In contrast, the black circles are our effect size and standard errors from the data generated in the random effect meta-analysis (RE) dataset. The red line is the average, true effect size (2).

```{r, fvsr, eval=TRUE, echo=FALSE}

```
