---
title: "Nuisance heterogeneity"
author: Patrice Pottier and Daniel W.A. Noble
date: '`r Sys.Date()`'
bibliography: ./bib/refs.bib
csl: ./bib/the-journal-of-experimental-biology.csl
output:
  bookdown::html_document2:
    css: style.css
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, tidy = TRUE)
options(digits=2)
```

```{r klippy, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
#install.packages("devtools")
remotes::install_github("rlesur/klippy")
klippy::klippy(tooltip_message = 'Click to Copy Code', tooltip_success = 'Done', position = 'right', color = "red")
```

# Introduction




``{r loadpackages, echo = TRUE, results = "hide", messages = FALSE, class.source='klippy'}
#install.packages("pacman")
#devtools::install_github("itchyshin/orchard_plot", subdir = "orchaRd", force = TRUE)
#devtools::install_github("daniel1noble/metaAidR")
pacman::p_load(devtools, png, jpeg, pander, cowplot, flextable, formatR, magick,
               tidyverse, metafor, orchaRd, here, osfr, magrittr, metaAidR)
```

asr_dat <- read.csv("https://osf.io/qn2af/download")


## References

<div id="refs"></div>

<br>

## Session Information

```{r sessioninfo, echo = FALSE, tidy=TRUE}
# Here, we'll create some R code
pander(sessionInfo(), locale = FALSE)
```

## [Back to Table of Contents](https://daniel1noble.github.io/meta-workshop/) {.hide}

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
