---
title: "Software"
author: Daniel W.A. Noble
date: '`r Sys.Date()`'
bibliography: ./bib/refs.bib
csl: ./bib/the-journal-of-experimental-biology.csl
output:
  bookdown::html_document2:
    css: style.css
    code_folding: show
    number_sections: no
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, tidy = TRUE)
options(digits=2)
```

```{r klippy, echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
#install.packages("devtools")
remotes::install_github("rlesur/klippy")
klippy::klippy(tooltip_message = 'Click to Copy Code', tooltip_success = 'Done', position = 'right', color = "red")

# Load packages
pacman::p_load(metafor, flextable, tidyverse, orchaRd, pander, mathjaxr, equatags, vembedr)

# To use mathjaxr you need to run equatags::mathjax_install()
```

# **Useful Software**

This page compiles some of the useful software tools that are out there for various steps of the meta-analysis workflow. We primarily collate software that we use and that fills important gaps. This is by no means comprehensive and there are many software packages that are underdevelopment. The [ESHhackathon webpage](https://www.eshackathon.org/projects.html) is a good place to look, but we list the ones we have found particularly useful.

### **Literature Searching, Screening Snowballing**

#### [litsearchr](https://elizagrames.github.io/litsearchr/)
```{r, echo=FALSE}
embed_url("https://www.youtube.com/watch?v=Z0GWTzl9OCE") %>% use_align("center")
```

#### [Revtools](https://onlinelibrary.wiley.com/doi/10.1002/jrsm.1374)
```{r, echo=FALSE}
embed_url("https://www.youtube.com/watch?v=WMpPCvBeILQ") %>% use_align("center")
```

#### [citationchaser](https://estech.shinyapps.io/citationchaser/) and see [paper](https://onlinelibrary.wiley.com/doi/10.1002/jrsm.1563?af=R)

#### Grey literature [greylitsearcher](https://estech.shinyapps.io/greylitsearcher/) 

### **Extracting Data from Figures** 

#### [metaDigitise](https://github.com/daniel1noble/metaDigitise) and [ShinyDigitise](https://github.com/EIvimeyCook/ShinyDigitise)
```{r, echo=FALSE}
embed_url("https://www.youtube.com/watch?v=2Q8TzgRSACM") %>% use_align("center")
```
#### [juicR](https://cran.r-project.org/web/packages/juicr/index.html)
```{r, echo=FALSE}
embed_url("https://www.youtube.com/watch?v=tiL-gZgN9Qk") %>% use_align("center")
```
